# (PART) STATISTICAL ANALYSIS {-}

# How to Perform Statistical Analysis in Python and R?

## Explanation

Statistical analysis helps us understand the characteristics of our dataset, identify patterns, and make data-driven decisions. In this section, we will cover basic statistical measures such as mean, median, variance, and correlation.

```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(
  echo  =TRUE,
  message  =FALSE,
  warning  =FALSE,
  cache  =FALSE,
  comment  =NA
)

if(!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)}
```

## Python Code


```python
import pandas as pd

# Load dataset
df = pd.read_csv("data/iris.csv")

# Summary statistics
summary_stats = df.describe()

# Calculate variance for numerical columns
variance = df.var(numeric_only=True)

# Calculate correlation between numerical variables
correlation = df.corr(numeric_only=True)

# Display results
print("Summary Statistics:\n", summary_stats)
print("\nVariance:\n", variance)
print("\nCorrelation:\n", correlation)
```

    Summary Statistics:
            sepal_length  sepal_width  petal_length  petal_width      target
    count    150.000000   150.000000    150.000000   150.000000  150.000000
    mean       5.843333     3.057333      3.758000     1.199333    1.000000
    std        0.828066     0.435866      1.765298     0.762238    0.819232
    min        4.300000     2.000000      1.000000     0.100000    0.000000
    25%        5.100000     2.800000      1.600000     0.300000    0.000000
    50%        5.800000     3.000000      4.350000     1.300000    1.000000
    75%        6.400000     3.300000      5.100000     1.800000    2.000000
    max        7.900000     4.400000      6.900000     2.500000    2.000000
    
    Variance:
     sepal_length    0.685694
    sepal_width     0.189979
    petal_length    3.116278
    petal_width     0.581006
    target          0.671141
    dtype: float64
    
    Correlation:
                   sepal_length  sepal_width  petal_length  petal_width    target
    sepal_length      1.000000    -0.117570      0.871754     0.817941  0.782561
    sepal_width      -0.117570     1.000000     -0.428440    -0.366126 -0.426658
    petal_length      0.871754    -0.428440      1.000000     0.962865  0.949035
    petal_width       0.817941    -0.366126      0.962865     1.000000  0.956547
    target            0.782561    -0.426658      0.949035     0.956547  1.000000


## R Code

```{r}
# Load dataset
df <- read.csv("data/iris.csv")

# Summary statistics
summary_stats <- summary(df)

# Calculate variance for numerical columns
variance <- apply(df[, 1:4], 2, var)

# Calculate correlation between numerical variables
correlation <- cor(df[, 1:4])

# Display results
print("Summary Statistics:")
print(summary_stats)
print("\nVariance:")
print(variance)
print("\nCorrelation:")
print(correlation)

```

# How to Calculate Skewness and Kurtosis in Python and R?

## Explanation

Skewness and kurtosis help us understand the distribution of data.  
- **Skewness** measures the asymmetry of the data distribution. A skewness of 0 indicates a perfectly symmetric distribution.  
- **Kurtosis** measures the "tailedness" of the distribution. A normal distribution has a kurtosis of 3. Values greater than 3 indicate heavy tails, while values less than 3 indicate light tails.

## Python Code




```python
import pandas as pd
from scipy.stats import skew, kurtosis

# Load dataset
df = pd.read_csv("data/iris.csv")

# Compute skewness
skewness = df.iloc[:, :-1].apply(skew)

# Compute kurtosis
kurt = df.iloc[:, :-1].apply(kurtosis)

# Display results
print("Skewness:\n", skewness)
print("\nKurtosis:\n", kurt)
```

    Skewness:
     sepal_length    0.311753
    sepal_width     0.315767
    petal_length   -0.272128
    petal_width    -0.101934
    target          0.000000
    dtype: float64
    
    Kurtosis:
     sepal_length   -0.573568
    sepal_width     0.180976
    petal_length   -1.395536
    petal_width    -1.336067
    target         -1.500000
    dtype: float64


## R Code

```{r}
# Check and load necessary libraries from CRAN mirror
if(!require(tidyverse)) install.packages("tidyverse", dependencies = TRUE, repos = "https://cloud.r-project.org/")
if(!require(e1071)) install.packages("e1071", dependencies = TRUE, repos = "https://cloud.r-project.org/")

library(tidyverse)
library(e1071)

# Load dataset
df <- read_csv("data/iris.csv")

# Compute skewness and kurtosis
skewness_values <- df %>%
  select(-species) %>%
  summarise(across(everything(), skewness))

kurtosis_values <- df %>%
  select(-species) %>%
  summarise(across(everything(), kurtosis))

# Display results
print("Skewness:")
print(skewness_values)

print("Kurtosis:")
print(kurtosis_values)
```

# How to Perform a t-test in Python and R?

## Explanation

**t-tests** are used to compare the means of two groups and determine whether they are significantly different from each other. In the iris dataset, we can compare the sepal length of two species to see if their means differ significantly.

There are different types of t-tests:

**Independent t-test**: Compares means between two independent groups.

**Paired t-test**: Compares means from the same group at different time points.

## Python Code

In Python, we use **scipy.stats.ttest_ind()** for an independent t-test.


```python
import pandas as pd
from scipy import stats

# Load dataset
df = pd.read_csv("data/iris.csv")

# Filter two species for comparison
setosa = df[df['species'] == 'setosa']['sepal_length']
versicolor = df[df['species'] == 'versicolor']['sepal_length']

# Perform independent t-test
t_stat, p_value = stats.ttest_ind(setosa, versicolor)

print(f"t-statistic: {t_stat}, p-value: {p_value}")
```

    t-statistic: -10.52098626754911, p-value: 8.985235037487079e-18


# How to compute the mean, median, and mode of a dataset?

## Explanation
- **Mean**: The average of all values in the dataset.
- **Median**: The middle value when the data is sorted.
- **Mode**: The value that appears most frequently in the dataset.

## Python Code



```python
import pandas as pd

# Load dataset
df = pd.read_csv("data/iris.csv")

# Compute mean, median, and mode
mean_values = df.drop(columns=["species"]).mean()
median_values = df.drop(columns=["species"]).median()
mode_values = df.drop(columns=["species"]).mode().iloc[0]

# Display results
print("Mean:\n")
print(mean_values)

print("\nMedian:\n")
print(median_values)

print("\nMode:\n")
print(mode_values)
```

    Mean:
    
    sepal_length    5.843333
    sepal_width     3.057333
    petal_length    3.758000
    petal_width     1.199333
    target          1.000000
    dtype: float64
    
    Median:
    
    sepal_length    5.80
    sepal_width     3.00
    petal_length    4.35
    petal_width     1.30
    target          1.00
    dtype: float64
    
    Mode:
    
    sepal_length    5.0
    sepal_width     3.0
    petal_length    1.4
    petal_width     0.2
    target          0.0
    Name: 0, dtype: float64


## R Code

```{r}
# Load necessary libraries
library(tidyverse)

# Load dataset
df <- read_csv("data/iris.csv")

# Compute mean, median, and mode
mean_values <- df %>%
  select(-species) %>%
  summarise(across(everything(), mean))

median_values <- df %>%
  select(-species) %>%
  summarise(across(everything(), median))

mode_values <- df %>%
  select(-species) %>%
  summarise(across(everything(), ~ names(sort(table(.), decreasing = TRUE))[1]))

# Display results
print("Mean:")
print(mean_values)

print("Median:")
print(median_values)

print("Mode:")
print(mode_values)

```

# What is the Difference Between an F-test and an ANOVA Test?

## Overview  
Both the **F-test** and **ANOVA** use the **F-statistic**, but they serve different purposes in statistical analysis.  

## Key Aspects  

| **Aspect**         | **F-test** (Variance Comparison)                   | **ANOVA** (Mean Comparison)                     |
|--------------------|---------------------------------------------------|-------------------------------------------------|
| **Purpose**       | Compares the variances of two groups.              | Compares the means of three or more groups.     |
| **Hypotheses**    | - \( H_0 \): Variances are equal. <br> - \( H_a \): Variances are different. | - \( H_0 \): All group means are equal. <br> - \( H_a \): At least one mean is different. |
| **When to Use?**  | Before a t-test, to check variance equality.        | When analyzing differences among multiple groups. |
| **Test Statistic** | \( F = \frac{\sigma_1^2}{\sigma_2^2} \)  (Ratio of variances) | \( F = \frac{\text{Between-group variance}}{\text{Within-group variance}} \) |
| **Python Function** | `levene()` or `bartlett()` from `scipy.stats`.    | `f_oneway()` from `scipy.stats` (for means). |
| **R Function**     | `var.test(group1, group2)`.                        | `aov(response ~ group, data = df)`. |

## Key Differences  

- The **F-test** is used to compare **variances** between two groups.  
- **ANOVA** is used to compare **means** among **three or more groups**.  
- The **F-test** is often used **before ANOVA** to check if the assumption of equal variances holds.  

## What If Variances Are Not Equal?  
If the assumption of equal variances is violated, consider using:  
- **Welch’s ANOVA**, which does not assume equal variances.  
- **Non-parametric tests**, such as the **Kruskal-Wallis test** (for comparing medians).  

---


# How to Perform an F-test in Python and R?

## Explanation  
An **F-test** is used to compare the variances of two independent groups. It helps determine if the groups have equal variances, which is important in statistical tests like t-tests and ANOVA.

The null hypothesis (\(H_0\)) assumes that the variances of the two groups are equal, while the alternative hypothesis (\(H_a\)) states that they are different.

The **F-statistic** for the F-test is calculated as:

\[
F = \frac{\text{variance of group 1}}{\text{variance of group 2}}
\]

If the p-value is small (typically \( p < 0.05 \)), we reject the null hypothesis and conclude that the variances are significantly different.

## Python Code


```python
import pandas as pd
from scipy.stats import levene

# Load dataset
df = pd.read_csv("data/iris.csv")

# Select two species for comparison
group1 = df[df["species"] == "setosa"]["sepal_length"]
group2 = df[df["species"] == "versicolor"]["sepal_length"]

# Perform F-test for variances (Levene's test)
f_stat, p_value = levene(group1, group2)

# Display results
print(f"F-statistic: {f_stat:.4f}")
print(f"P-value: {p_value:.4f}")

# Interpretation
if p_value < 0.05:
    print("Reject H0: The variances of Setosa and Versicolor are significantly different.")
else:
    print("Fail to reject H0: No significant difference in variances.")
```

    F-statistic: 8.1727
    P-value: 0.0052
    Reject H0: The variances of Setosa and Versicolor are significantly different.


## R Code

```{r}
# Load necessary package
if (!require(car)) install.packages("car", repos = "https://cloud.r-project.org")
library(car)

# Load dataset
df <- read.csv("data/iris.csv")

# Subset data for two species
group1 <- df[df$species == "setosa", "sepal_length"]
group2 <- df[df$species == "versicolor", "sepal_length"]

# Perform F-test for variances (Levene's test)
var_test_result <- var.test(group1, group2)
print(var_test_result)
```

# How to Perform an ANOVA Test in Python and R?

## Explanation

An ANOVA (Analysis of Variance) test is used to compare the means of three or more groups. It helps to determine whether there are any statistically significant differences between the means of the groups.

The null hypothesis ((H_0)) assumes that all group means are equal, while the alternative hypothesis ((H_a)) states that at least one mean is different.

The F-statistic in ANOVA is calculated as:

\[
F = \frac{\text{Between-group variance}}{\text{Within-group variance}}
\]

If the p-value is small (typically ( p < 0.05 )), we reject the null hypothesis and conclude that at least one group mean is significantly different.

## Python Code


```python
import pandas as pd
from scipy.stats import f_oneway

# Load dataset
df = pd.read_csv("data/iris.csv")

# Select groups for comparison
group1 = df[df["species"] == "setosa"]["sepal_length"]
group2 = df[df["species"] == "versicolor"]["sepal_length"]
group3 = df[df["species"] == "virginica"]["sepal_length"]

# Perform ANOVA
f_stat, p_value = f_oneway(group1, group2, group3)

# Display results
print(f"F-statistic: {f_stat:.4f}")
print(f"P-value: {p_value:.4f}")

# Interpretation
if p_value < 0.05:
    print("Reject H0: At least one group mean is significantly different.")
else:
    print("Fail to reject H0: No significant difference in group means.")
```

    F-statistic: 119.2645
    P-value: 0.0000
    Reject H0: At least one group mean is significantly different.


## R Code 

```{r}
# Load dataset
df <- read.csv("data/iris.csv")

# Perform ANOVA (sepal length by species)
anova_result <- aov(sepal_length ~ species, data = df)
summary(anova_result)
```

# How to Perform a Chi-Square Test in Python and R?

## Explanation  
A **Chi-Square Test** is used to determine whether there is a significant association between two categorical variables. It compares the observed frequencies in each category to the frequencies we would expect if the variables were independent.

The **Chi-Square Statistic** is calculated as:

\[
\chi^2 = \sum \frac{(O_i - E_i)^2}{E_i}
\]

Where:
- \(O_i\) is the observed frequency in category \(i\),
- \(E_i\) is the expected frequency in category \(i\).

The null hypothesis (\(H_0\)) assumes that there is no association between the variables (i.e., the variables are independent), while the alternative hypothesis (\(H_a\)) states that there is an association between them.

If the p-value is small (typically \( p < 0.05 \)), we reject the null hypothesis and conclude that there is a significant association between the variables.

## Python Code 
- **Chi-Square Test** - Association Between Categorical Variables



```python
import pandas as pd
from scipy.stats import chi2_contingency

# Load dataset
df = pd.read_csv("data/iris.csv")

# Create a contingency table for 'species' and 'sepal_width' (categorical grouping)
contingency_table = pd.crosstab(df['species'], pd.cut(df['sepal_width'], bins=3))

# Perform Chi-Square Test
chi2_stat, p_value, dof, expected = chi2_contingency(contingency_table)

# Display results
print(f"Chi-Square Statistic: {chi2_stat:.4f}")
print(f"P-value: {p_value:.4f}")
print(f"Degrees of Freedom: {dof}")
print(f"Expected Frequencies: \n{expected}")

# Interpretation
if p_value < 0.05:
    print("Reject H0: There is a significant association between species and sepal width.")
else:
    print("Fail to reject H0: No significant association between species and sepal width.")
```

    Chi-Square Statistic: 45.1247
    P-value: 0.0000
    Degrees of Freedom: 4
    Expected Frequencies: 
    [[15.66666667 29.33333333  5.        ]
     [15.66666667 29.33333333  5.        ]
     [15.66666667 29.33333333  5.        ]]
    Reject H0: There is a significant association between species and sepal width.


## R Code
Chi-Square Test - Association Between Categorical Variables

```{r}
# Load dataset
df <- read.csv("data/iris.csv")

# Create a contingency table for 'species' and 'sepal_width' (categorical grouping)
df$sepal_width_cat <- cut(df$sepal_width, breaks = 3)
contingency_table <- table(df$species, df$sepal_width_cat)

# Perform Chi-Square Test
chi2_result <- chisq.test(contingency_table)

# Display results
print(chi2_result)

# Interpretation
if (chi2_result$p.value < 0.05) {
    print("Reject H0: There is a significant association between species and sepal width.")
} else {
    print("Fail to reject H0: No significant association between species and sepal width.")
}
```

# How to Perform a Pearson Correlation Test in Python and R?

## Explanation  
The **Pearson Correlation Test** is used to determine the linear relationship between two continuous variables. It measures the strength and direction of the relationship, with a correlation coefficient (\(r\)) ranging from -1 to 1:
- \( r = 1 \) indicates a perfect positive linear relationship.
- \( r = -1 \) indicates a perfect negative linear relationship.
- \( r = 0 \) indicates no linear relationship.

The null hypothesis (\(H_0\)) assumes that there is no linear correlation between the two variables, while the alternative hypothesis (\(H_a\)) states that there is a linear correlation.

If the p-value is small (typically \( p < 0.05 \)), we reject the null hypothesis and conclude that there is a significant linear relationship between the two variables.

## Python Code



```python
import pandas as pd
from scipy.stats import pearsonr

# Load dataset
df = pd.read_csv("data/iris.csv")

# Select two variables for correlation test
x = df['sepal_length']
y = df['sepal_width']

# Perform Pearson Correlation Test
corr_coefficient, p_value = pearsonr(x, y)

# Display results
print(f"Pearson Correlation Coefficient: {corr_coefficient:.4f}")
print(f"P-value: {p_value:.4f}")

# Interpretation
if p_value < 0.05:
    print("Reject H0: There is a significant linear correlation between sepal length and sepal width.")
else:
    print("Fail to reject H0: No significant linear correlation between sepal length and sepal width.")
```

    Pearson Correlation Coefficient: -0.1176
    P-value: 0.1519
    Fail to reject H0: No significant linear correlation between sepal length and sepal width.


## R Code

```{r}
# Load dataset
df <- read.csv("data/iris.csv")

# Select two variables for correlation test
x <- df$sepal_length
y <- df$sepal_width

# Perform Pearson Correlation Test
cor_result <- cor.test(x, y)

# Display results
print(cor_result)

# Interpretation
if (cor_result$p.value < 0.05) {
    print("Reject H0: There is a significant linear correlation between sepal length and sepal width.")
} else {
    print("Fail to reject H0: No significant linear correlation between sepal length and sepal width.")
}
```

# How to Perform a Simple Linear Regression in Python and R?

## Explanation  
**Simple Linear Regression** is used to model the relationship between a dependent variable (\(Y\)) and an independent variable (\(X\)) by fitting a linear equation to observed data. The goal is to find the best-fitting line, represented by the equation:

\[
Y = \beta_0 + \beta_1 X + \epsilon
\]

Where:
- \(Y\) is the dependent variable,
- \(X\) is the independent variable,
- \(\beta_0\) is the intercept,
- \(\beta_1\) is the slope,
- \(\epsilon\) is the error term.

The null hypothesis (\(H_0\)) assumes that the slope of the regression line is zero, meaning there is no linear relationship between \(X\) and \(Y\). The alternative hypothesis (\(H_a\)) states that the slope is non-zero, indicating a significant relationship.

If the p-value is small (typically \( p < 0.05 \)), we reject the null hypothesis and conclude that there is a significant relationship between \(X\) and \(Y\).

## Python Code
Simple Linear Regression




```python
import pandas as pd
import statsmodels.api as sm

# Load dataset
df = pd.read_csv("data/iris.csv")

# Select independent and dependent variables
X = df['sepal_length']
y = df['sepal_width']

# Add a constant (intercept) to the independent variable
X = sm.add_constant(X)

# Perform linear regression
model = sm.OLS(y, X).fit()

# Display results
print(model.summary())

# Interpretation
if model.pvalues[1] < 0.05:
    print("Reject H0: There is a significant linear relationship between sepal length and sepal width.")
else:
    print("Fail to reject H0: No significant linear relationship between sepal length and sepal width.")
```

                                OLS Regression Results                            
    ==============================================================================
    Dep. Variable:            sepal_width   R-squared:                       0.014
    Model:                            OLS   Adj. R-squared:                  0.007
    Method:                 Least Squares   F-statistic:                     2.074
    Date:                Tue, 18 Mar 2025   Prob (F-statistic):              0.152
    Time:                        13:06:35   Log-Likelihood:                -86.732
    No. Observations:                 150   AIC:                             177.5
    Df Residuals:                     148   BIC:                             183.5
    Df Model:                           1                                         
    Covariance Type:            nonrobust                                         
    ================================================================================
                       coef    std err          t      P>|t|      [0.025      0.975]
    --------------------------------------------------------------------------------
    const            3.4189      0.254     13.484      0.000       2.918       3.920
    sepal_length    -0.0619      0.043     -1.440      0.152      -0.147       0.023
    ==============================================================================
    Omnibus:                        2.474   Durbin-Watson:                   1.263
    Prob(Omnibus):                  0.290   Jarque-Bera (JB):                1.994
    Skew:                           0.243   Prob(JB):                        0.369
    Kurtosis:                       3.288   Cond. No.                         43.4
    ==============================================================================
    
    Notes:
    [1] Standard Errors assume that the covariance matrix of the errors is correctly specified.
    Fail to reject H0: No significant linear relationship between sepal length and sepal width.


## R Code
Simple Linear Regression


```{r}
# Load dataset
df <- read.csv("data/iris.csv")

# Select independent and dependent variables
X <- df$sepal_length
y <- df$sepal_width

# Perform linear regression
model <- lm(y ~ X)

# Display results
summary(model)

# Interpretation
if (summary(model)$coefficients[2,4] < 0.05) {
    print("Reject H0: There is a significant linear relationship between sepal length and sepal width.")
} else {
    print("Fail to reject H0: No significant linear relationship between sepal length and sepal width.")
}
```

# How to Perform Multiple Linear Regression in Python and R?

## Explanation

**Multiple Linear Regression** is an extension of simple linear regression that models the relationship between a dependent variable (\(Y\)) and two or more independent variables (\(X_1, X_2, \dots, X_n\)). The equation for multiple linear regression is:

\[
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_n X_n + \epsilon
\]

Where:
- \(Y\) is the dependent variable,
- \(X_1, X_2, \dots, X_n\) are the independent variables,
- \(\beta_0\) is the intercept,
- \(\beta_1, \beta_2, \dots, \beta_n\) are the coefficients (slopes) for the predictors,
- \(\epsilon\) is the error term.

The null hypothesis (\(H_0\)) assumes that all regression coefficients are zero, meaning no relationship exists between the predictors and the dependent variable. The alternative hypothesis (\(H_a\)) suggests that at least one coefficient is non-zero, indicating a significant relationship.

If the p-value for a coefficient is small (typically \(p < 0.05\)), we reject the null hypothesis for that predictor and conclude that it significantly affects \(Y\).

## Python Code



```python
import pandas as pd
import statsmodels.api as sm

# Load dataset (you can use any dataset here)
df = pd.read_csv("data/iris.csv")

# Select independent variables (predictors)
X = df[['sepal_length', 'sepal_width', 'petal_length', 'petal_width']]  # Predictors

# Add a constant (intercept) to the model
X = sm.add_constant(X)

# Dependent variable (response)
y = df['sepal_length']  # Example of using sepal length as the dependent variable

# Fit the model
model = sm.OLS(y, X).fit()

# Display the results
print(model.summary())
```

                                OLS Regression Results                            
    ==============================================================================
    Dep. Variable:           sepal_length   R-squared:                       1.000
    Model:                            OLS   Adj. R-squared:                  1.000
    Method:                 Least Squares   F-statistic:                 9.587e+29
    Date:                Tue, 18 Mar 2025   Prob (F-statistic):               0.00
    Time:                        13:32:45   Log-Likelihood:                 4724.3
    No. Observations:                 150   AIC:                            -9439.
    Df Residuals:                     145   BIC:                            -9424.
    Df Model:                           4                                         
    Covariance Type:            nonrobust                                         
    ================================================================================
                       coef    std err          t      P>|t|      [0.025      0.975]
    --------------------------------------------------------------------------------
    const        -9.104e-15   4.83e-15     -1.887      0.061   -1.86e-14    4.34e-16
    sepal_length     1.0000   1.36e-15   7.36e+14      0.000       1.000       1.000
    sepal_width    1.11e-15   1.41e-15      0.790      0.431   -1.67e-15    3.89e-15
    petal_length  2.776e-16   1.34e-15      0.207      0.836   -2.37e-15    2.92e-15
    petal_width  -1.554e-15   2.23e-15     -0.698      0.486   -5.95e-15    2.84e-15
    ==============================================================================
    Omnibus:                        5.539   Durbin-Watson:                   0.030
    Prob(Omnibus):                  0.063   Jarque-Bera (JB):                3.049
    Skew:                           0.099   Prob(JB):                        0.218
    Kurtosis:                       2.330   Cond. No.                         91.9
    ==============================================================================
    
    Notes:
    [1] Standard Errors assume that the covariance matrix of the errors is correctly specified.


## R Code
Multiple Linear Regression

```{r}
# Check if 'caret' is installed, if not, install it
if (!require(caret)) {
  # Set CRAN mirror
  options(repos = c(CRAN = "https://cran.rstudio.com/"))
  install.packages("caret")
  library(caret)
}
# Load dataset (use any dataset available)
df <- read.csv("data/iris.csv")

# Create training and test sets
set.seed(123)  # For reproducibility
trainIndex <- createDataPartition(df$sepal_length, p = 0.7, list = FALSE)
trainData <- df[trainIndex, ]
testData <- df[-trainIndex, ]

# Train the linear regression model
model <- train(sepal_length ~ sepal_width + petal_length + petal_width,
               data = trainData,
               method = "lm")

# Display the model details
print(model)

# Predict on the test set
predictions <- predict(model, testData)

# Display predictions and actual values
predictions
testData$sepal_length

# Evaluate the model using RMSE (Root Mean Squared Error)
rmse <- sqrt(mean((predictions - testData$sepal_length)^2))
cat("RMSE: ", rmse)
```

# How to Perform Logistic Regression in Python and R?

## Explanation

**Logistic Regression** is a statistical method used for binary classification. It models the relationship between a dependent variable (binary outcome, e.g., 0 or 1) and one or more independent variables. The logistic regression equation is:

\[
\log\left(\frac{p}{1 - p}\right) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_n X_n
\]

Where:
- \( p \) is the probability of the dependent event occurring (i.e., the probability that the output is 1),
- \( X_1, X_2, \dots, X_n \) are the independent variables (predictors),
- \( \beta_0 \) is the intercept,
- \( \beta_1, \beta_2, \dots, \beta_n \) are the coefficients (slopes) for the predictors.

The model estimates the odds of the event occurring by taking the log of the odds ratio. The predicted probabilities are obtained using the logistic function, which is:

\[
p = \frac{1}{1 + e^{-z}}
\]

Where \( z = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_n X_n \).

If the p-value of a coefficient is small (typically \( p < 0.05 \)), we reject the null hypothesis and conclude that the predictor significantly influences the outcome.

## Python Code




```python
import pandas as pd
from sklearn.model_selection import train_test_split
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import accuracy_score, confusion_matrix

# Load dataset (you can use any binary classification dataset)
df = pd.read_csv("data/iris.csv")

# Select independent variables (predictors)
X = df[['sepal_length', 'sepal_width', 'petal_length', 'petal_width']]  # Predictors

# Create a binary dependent variable (response)
df['is_setosa'] = (df['species'] == 'setosa').astype(int)  # Convert 'setosa' to 1, others to 0
y = df['is_setosa']  # Target variable

# Split data into training and testing sets
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# Fit the logistic regression model
model = LogisticRegression()
model.fit(X_train, y_train)

# Make predictions
y_pred = model.predict(X_test)

# Evaluate the model
print("Accuracy:", accuracy_score(y_test, y_pred))
print("Confusion Matrix:\n", confusion_matrix(y_test, y_pred))
```

    Accuracy: 1.0
    Confusion Matrix:
     [[26  0]
     [ 0 19]]


## R Code

```{r, eval=FALSE}
# Load required library
if (!require(caret)) {
  install.packages("caret")
  library(caret)
}

# Load dataset (use any binary classification dataset)
df <- read.csv("data/iris.csv")

# Create a binary dependent variable (response)
df$is_setosa <- ifelse(df$species == 'setosa', 1, 0)  # Convert 'setosa' to 1, others to 0

# Create training and test sets
set.seed(123)  # For reproducibility
trainIndex <- createDataPartition(df$is_setosa, p = 0.7, list = FALSE)
trainData <- df[trainIndex, ]
testData <- df[-trainIndex, ]

# Train the logistic regression model
model <- train(is_setosa ~ sepal_length + sepal_width + petal_length + petal_width,
               data = trainData,
               method = "glm",
               family = "binomial",
               trControl = trainControl(method = "cv", number = 10))

# Display the model details
print(model)

# Predict the probabilities using the logistic regression model
pred_probs <- predict(model, testData, type = "prob")[, 2]  # Get the probability of class 1

# Convert probabilities to class predictions (0 or 1)
pred_class <- ifelse(pred_probs > 0.5, 1, 0)

# Evaluate the model using confusion matrix
conf_matrix <- confusionMatrix(factor(pred_class), factor(testData$is_setosa))

# Print confusion matrix
print(conf_matrix)
```

# How to Perform Decision Tree Classification in Python and R?

## Explanation

**Decision Trees** are a non-linear model used for both classification and regression. They split the dataset into subsets based on the most significant features, creating a tree-like structure where each internal node represents a feature or attribute, and each leaf node represents a decision or class label.

The decision tree algorithm works by selecting the feature that best splits the data at each node based on certain criteria (e.g., Gini impurity, entropy for classification, and variance reduction for regression). 

- **Classification Trees**: Used when the target variable is categorical. For example, in a binary classification problem, the decision tree can predict class 0 or class 1 based on the input features.
- **Regression Trees**: Used when the target variable is continuous.

The tree is built recursively by selecting the best splits at each node and stopping when a stopping criterion (e.g., maximum depth, minimum samples per leaf) is met.

## Python Code



```python
import pandas as pd
from sklearn.model_selection import train_test_split
from sklearn.tree import DecisionTreeClassifier
from sklearn.metrics import accuracy_score, confusion_matrix

# Load dataset (you can use any dataset)
df = pd.read_csv("data/iris.csv")

# Select independent variables (predictors)
X = df[['sepal_length', 'sepal_width', 'petal_length', 'petal_width']]  # Features

# Select target variable (species)
y = df['species']  # Target

# Split data into training and testing sets
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# Fit the decision tree classifier
model = DecisionTreeClassifier(random_state=42)
model.fit(X_train, y_train)

# Make predictions
y_pred = model.predict(X_test)

# Evaluate the model
print("Accuracy:", accuracy_score(y_test, y_pred))
print("Confusion Matrix:\n", confusion_matrix(y_test, y_pred))
```

    Accuracy: 1.0
    Confusion Matrix:
     [[19  0  0]
     [ 0 13  0]
     [ 0  0 13]]


## R Code

```{r, eval=FALSE}
# Load required library
library(caret)

# Load dataset (use any dataset)
df <- read.csv("data/iris.csv")

# Create training and test sets
set.seed(123)  # For reproducibility
trainIndex <- createDataPartition(df$species, p = 0.7, list = FALSE)
trainData <- df[trainIndex, ]
testData <- df[-trainIndex, ]

# Train the model using decision tree classifier
model <- train(species ~ sepal_length + sepal_width + petal_length + petal_width,
               data = trainData,
               method = "rpart",
               trControl = trainControl(method = "cv", number = 10))

# Display the model details
print(model)

# Predict using the decision tree model
pred <- predict(model, testData)

# Evaluate the model using confusion matrix
confusionMatrix(pred, testData$species)
```

# How to Perform Random Forest Classification in Python and R?

## Explanation

**Random Forest** is an ensemble learning method that combines multiple decision trees to improve classification accuracy. Instead of relying on a single decision tree, random forest aggregates the predictions of many trees, reducing overfitting and improving generalization.

- **Bagging (Bootstrap Aggregating)**: Random Forest uses bagging, which means training multiple models (decision trees) on random subsets of the data. Each tree is trained on a different random sample, and the final prediction is made by averaging (for regression) or majority voting (for classification) of all the trees' predictions.
- **Random Feature Selection**: At each split in the decision tree, a random subset of features is selected, ensuring that trees are diverse and reducing the correlation between them.

The main advantages of random forest are its robustness, the ability to handle a large number of features, and its capacity to deal with overfitting.

## Python Code




```python
import pandas as pd
from sklearn.model_selection import train_test_split
from sklearn.ensemble import RandomForestClassifier
from sklearn.metrics import accuracy_score, confusion_matrix

# Load dataset (you can use any dataset)
df = pd.read_csv("data/iris.csv")

# Select independent variables (predictors)
X = df[['sepal_length', 'sepal_width', 'petal_length', 'petal_width']]  # Features

# Select target variable (species)
y = df['species']  # Target

# Split data into training and testing sets
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# Fit the random forest classifier
model = RandomForestClassifier(n_estimators=100, random_state=42)
model.fit(X_train, y_train)

# Make predictions
y_pred = model.predict(X_test)

# Evaluate the model
print("Accuracy:", accuracy_score(y_test, y_pred))
print("Confusion Matrix:\n", confusion_matrix(y_test, y_pred))
```

    Accuracy: 1.0
    Confusion Matrix:
     [[19  0  0]
     [ 0 13  0]
     [ 0  0 13]]


## R Code

```{r, eval=FALSE}
# Load required library
library(caret)

# Load dataset (use any dataset)
df <- read.csv("data/iris.csv")

# Create training and test sets
set.seed(123)  # For reproducibility
trainIndex <- createDataPartition(df$species, p = 0.7, list = FALSE)
trainData <- df[trainIndex, ]
testData <- df[-trainIndex, ]

# Train the model using Random Forest
model <- train(species ~ sepal_length + sepal_width + petal_length + petal_width,
               data = trainData,
               method = "rf",
               trControl = trainControl(method = "cv", number = 10))

# Display the model details
print(model)

# Predict using the random forest model
pred <- predict(model, testData)

# Evaluate the model using confusion matrix
confusionMatrix(pred, testData$species)
```

# How to Perform Support Vector Machine (SVM) Classification in Python and R?

## Explanation

**Support Vector Machine (SVM)** is a powerful supervised learning algorithm that can be used for both classification and regression tasks. It works by finding the hyperplane that best separates data points of different classes in a high-dimensional space.

- **Linear SVM**: Finds a straight line or hyperplane that divides the classes.
- **Non-linear SVM**: Uses kernel functions (like Radial Basis Function (RBF)) to transform the data into higher dimensions to make it linearly separable.

The main objective of SVM is to maximize the margin between the two classes. The margin is the distance between the hyperplane and the closest data points from either class, known as support vectors.

The SVM classifier works well for both linear and non-linear classification problems.

## Python Code




```python
import pandas as pd
from sklearn.model_selection import train_test_split
from sklearn.svm import SVC
from sklearn.metrics import accuracy_score, confusion_matrix

# Load dataset (you can use any dataset)
df = pd.read_csv("data/iris.csv")

# Select independent variables (predictors)
X = df[['sepal_length', 'sepal_width', 'petal_length', 'petal_width']]  # Features

# Select target variable (species)
y = df['species']  # Target

# Split data into training and testing sets
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# Fit the SVM classifier
model = SVC(kernel='linear', random_state=42)
model.fit(X_train, y_train)

# Make predictions
y_pred = model.predict(X_test)

# Evaluate the model
print("Accuracy:", accuracy_score(y_test, y_pred))
print("Confusion Matrix:\n", confusion_matrix(y_test, y_pred))
```

    Accuracy: 1.0
    Confusion Matrix:
     [[19  0  0]
     [ 0 13  0]
     [ 0  0 13]]


## R Code

```{r, eval=FALSE}
# Load required library
library(caret)

# Load dataset (use any dataset)
df <- read.csv("data/iris.csv")

# Create training and test sets
set.seed(123)  # For reproducibility
trainIndex <- createDataPartition(df$species, p = 0.7, list = FALSE)
trainData <- df[trainIndex, ]
testData <- df[-trainIndex, ]

# Train the model using SVM with a linear kernel
model <- train(species ~ sepal_length + sepal_width + petal_length + petal_width,
               data = trainData,
               method = "svmLinear",
               trControl = trainControl(method = "cv", number = 10))

# Display the model details
print(model)

# Predict using the SVM model
pred <- predict(model, testData)

# Evaluate the model using confusion matrix
confusionMatrix(pred, testData$species)
```
